---
title: "LOVAC Statistics"
output: html_document
---
# Set Up
```{r}
library(tidyverse)
library(readr)
library(scales)
library(ggrepel)
library(patchwork)
library(gt)
library(knitr)
library(here)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(color = "grey40", size = 11),
      plot.caption  = element_text(color = "grey50", size = 9),
      panel.grid.minor = element_blank()
    )
)

MARSEILLE_ARR <- sprintf("132%02d", 1:16)   # 13201 … 13216
```

LOVAC - Descriptive Analysis: Local property-tax strategy in
Marseille

Data: Logements Vacants du Parc Privé (LOVAC), CEREMA /data.gouv.fr

NOTE ON MARSEILLE: Marseille is NOT stored under a single code in LOVAC.
It appears as 16 arrondissements with INSEE codes 13201 to 13216. All
analysis requires aggregating these 16 rows before comparing Marseille
to other cities. Similarly, Paris (75056 split into 75101–75120) and
Lyon (69123 split into 69381–69389) may need the same treatment
depending on the LOVAC edition.

# Data Set
```{r}
lovac_peers <- read.csv(here("cache", "lovac_peers.csv")) %>% 
  mutate(
    CODGEO_25 = as.character(CODGEO_25),
    year = as.character(year)
  )

lovac_long <- read.csv(here("cache", "lovac_long.csv" )) %>% 
    mutate(
    CODGEO_25 = as.character(CODGEO_25),
    year = as.character(year)
  )
```


# Descriptive analalisis LOVAC
## Summary statictics
```{r}
latest_year <- "2024" # 2025 is incomplete
 
summary_tbl <- lovac_peers %>%
  filter(year == latest_year) %>%
  arrange(desc(tx_lv_sup2)) %>%
  select(LIBGEO_25, nb_lp, nb_lv, nb_lv_sup2, nb_lv_inf2,
         tx_lv, tx_lv_sup2, tx_lv_inf2, share_structural)
 
t <- summary_tbl %>%
  gt() %>%
  tab_header(
    title    = "LOVAC — Vacant private dwellings, peer cities",
    subtitle = glue::glue("Year {latest_year} · Marseille aggregated from 16 arrondissements")
  ) %>%
  fmt_number(columns = c(nb_lp, nb_lv, nb_lv_sup2, nb_lv_inf2), decimals = 0) %>%
  fmt_percent(columns = c(tx_lv, tx_lv_sup2, tx_lv_inf2, share_structural),
              decimals = 1) %>%
  cols_label(
    LIBGEO_25        = "City",
    nb_lp            = "Private stock",
    nb_lv            = "Vacant (total)",
    nb_lv_sup2       = "Structural (2+ yr)",
    nb_lv_inf2       = "Frictional (<2 yr)",
    tx_lv            = "Vacancy rate",
    tx_lv_sup2       = "Structural rate",
    tx_lv_inf2       = "Frictional rate",
    share_structural = "% structural"
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#DBEAFE"), cell_text(weight = "bold")),
    locations = cells_body(rows = LIBGEO_25 == "Marseille")
  ) %>%
  tab_footnote(
    footnote = "Marseille: sum of arrondissements 13201–13216. Paris: 75101–75120. Lyon: 69381–69389.",
    locations = cells_column_labels(columns = LIBGEO_25)
  ) %>%
  tab_source_note("Source: LOVAC open data, CEREMA / DGFiP")
t
```

### Export Table
```{r}
gtsave(t, here("output", "LOVAC —Vacant private dwellings.png"))
```

## MARSEILLE LONGITUDINAL ANALYSIS
```{r}
marseille <- lovac_peers %>% filter(CODGEO_25 == "13055")
 
# --- 5.1 Stock by vacancy type (stacked bar) ---------------------------------
p_stock <- marseille %>%
  select(year, nb_lv_sup2, nb_lv_inf2) %>%
  pivot_longer(-year, names_to = "type", values_to = "n") %>%
  mutate(type = recode(type,
    nb_lv_sup2 = "Structural (2+ years)",
    nb_lv_inf2 = "Frictional (< 2 years)"
  )) %>%
  ggplot(aes(x = year, y = n, fill = type)) +
  geom_col(position = "stack", width = 0.65) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_manual(values = c("#1565C0", "#90CAF9")) +
  labs(
    title    = "Marseille — Vacant private dwellings by type",
    subtitle = "Aggregated from 16 arrondissements (13201–13216)",
    x = NULL, y = "Dwellings", fill = NULL,
    caption  = "Source: LOVAC open data"
  ) +
  theme(legend.position = "bottom")
 
# --- 5.2 Vacancy rates over time --------------------------------------------
p_rate <- marseille %>%
  select(year, tx_lv, tx_lv_sup2, tx_lv_inf2) %>%
  pivot_longer(-year, names_to = "indicator", values_to = "rate") %>%
  mutate(indicator = recode(indicator,
    tx_lv      = "Total",
    tx_lv_sup2 = "Structural (2+ yr)",
    tx_lv_inf2 = "Frictional (< 2 yr)"
  )) %>%
  ggplot(aes(x = year, y = rate, color = indicator, group = indicator)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1), limits = c(0, NA)) +
  scale_color_manual(values = c("#1565C0", "#42A5F5", "#90CAF9")) +
  labs(
    title = "Marseille — Vacancy rate trends",
    x = NULL, y = "Rate", color = NULL,
    caption = "Source: LOVAC open data"
  ) +
  theme(legend.position = "bottom")
 
# --- 5.3 Arrondissement breakdown (latest year) ------------------------------
marseille_arr_latest <- lovac_long %>%
  filter(CODGEO_25 %in% MARSEILLE_ARR, year == latest_year)  %>% 
  mutate(
    tx_lv      = nb_lv / nb_lp,
    tx_lv_sup2 = nb_lv_sup2 / nb_lp,
    arr_label  = paste0(substr(CODGEO_25, 4, 5), "e"),  # "01e", "02e", etc.
    arr_label  = fct_reorder(arr_label, tx_lv_sup2)
  )
 
p_arr <- marseille_arr_latest %>%
  ggplot(aes(x = arr_label, y = tx_lv_sup2,
             fill = tx_lv_sup2)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = percent(tx_lv_sup2, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_gradient(low = "#DBEAFE", high = "#1E3A8A", guide = "none") +
  coord_flip() +
  labs(
    title    = "Marseille — Structural vacancy by arrondissement",
    subtitle = glue::glue("Year {latest_year}"),
    x = NULL, y = "Structural vacancy rate",
    caption  = "Source: LOVAC open data"
  )
 
# Save combined Marseille panel
(p_stock / p_rate) | p_arr
ggsave(here("output" ,"marseille_vacancy_analysis.png"), width = 14, height = 8, dpi = 150)
```

# Peer comparison
```{r}
p_peers_bar <- lovac_peers %>%
  filter(year == latest_year) %>%
  mutate(
    LIBGEO_25    = fct_reorder(LIBGEO_25, tx_lv_sup2),
    highlight = (CODGEO_25 == "13055")
  ) %>%
  ggplot(aes(x = LIBGEO_25, y = tx_lv_sup2, fill = highlight)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = percent(tx_lv_sup2, accuracy = 0.1)),
            hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1),
                     expand  = expansion(mult = c(0, 0.14))) +
  scale_fill_manual(values = c("grey75", "#1565C0"), guide = "none") +
  coord_flip() +
  labs(
    title    = "Structural vacancy rate — peer comparison",
    subtitle = glue::glue("Year {latest_year}"),
    x = NULL, y = "Structural vacancy rate (2+ years)",
    caption  = "Source: LOVAC open data · Marseille, Paris, Lyon aggregated from arrondissements"
  )
 
# --- Multi-year trends ---------------------------------------------------
p_trends <- lovac_peers %>%
  mutate(highlight = (CODGEO_25 == "13055")) %>%
  ggplot(aes(x = year, y = tx_lv_sup2,
             group = LIBGEO_25, color = highlight,
             linewidth = highlight)) +
  geom_line() +
  geom_point(size = 2) +
  geom_text_repel(
    data = \(d) filter(d, year == max(as.numeric(year))-1),
    aes(label = LIBGEO_25), size = 3.2, nudge_x = 0.15, direction = "y"
  ) +
  scale_color_manual(values = c("grey70", "#1565C0"), guide = "none") +
  scale_linewidth_manual(values = c(0.7, 1.6), guide = "none") +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  scale_x_discrete(breaks = unique(lovac_peers$year)) +
  labs(
    title    = "Structural vacancy trends — Marseille vs. peers",
    subtitle = "Marseille highlighted in blue",
    x = NULL, y = "Structural vacancy rate",
    caption  = "Source: LOVAC open data"
  )
 
p_peers_bar / p_trends
ggsave(here("output" ,"peer_comparison.png"), width = 10, height = 10, dpi = 150)
```