---
title: "Potential rev"
output: html_document
---
# Set Up
```{r}
library(tidyverse)
library(readr)
library(scales)
library(ggrepel)
library(patchwork)
library(gt)
library(knitr)
library(here)
library(glue)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(color = "grey40", size = 11),
      plot.caption  = element_text(color = "grey50", size = 9),
      panel.grid.minor = element_blank()
    )
)
```

# Data Set
```{r}
rei_lovac <- read.csv(here("cache", "rei_lovac.csv"))
```

# Descriptive analalisis REI-LOVAC
## Summary statictics
```{r}
summary_table <- rei_lovac %>%
  select(annee, base_nette, taux_vote, taux_metro, taux_total,
         montant_reel, nb_articles, base_par_article,
         nb_lv_sup2, tx_lv_sup2,
         recettes_potentielles, gain_25pct) %>%
  mutate(
    base_nette            = round(base_nette / 1e6, 1),
    montant_reel          = round(montant_reel / 1e6, 1),
    base_par_article      = round(base_par_article, 0),
    nb_lv_sup2            = round(nb_lv_sup2, 0),
    tx_lv_sup2            = round(tx_lv_sup2 * 100, 2),
    recettes_potentielles = round(recettes_potentielles / 1e6, 1),
    gain_25pct            = round(gain_25pct / 1e6, 1)
  ) %>%
  rename(
    Year                    = annee,
    `Base (M€)`             = base_nette,
    `Rate commune (%)`      = taux_vote,
    `Rate métropole (%)`    = taux_metro,
    `Total rate (%)`        = taux_total,
    `Revenue (M€)`          = montant_reel,
    `N articles`            = nb_articles,
    `Base/article (€)`      = base_par_article,
    `Struct. vacant dwell.` = nb_lv_sup2,
    `Struct. vacancy rate`  = tx_lv_sup2,
    `Max foregone rev (M€)` = recettes_potentielles,
    `Gain 25% mobil. (M€)`  = gain_25pct
  )
```

## Table construction
```{r}
t <- summary_table %>%
  gt() %>%
  tab_header(
    title    = "LOVAC — Fiscal summary",
    subtitle = glue("Period covered: {min(summary_table$Year)}–{max(summary_table$Year)}")
  ) %>%
  fmt_number(
    columns = c(`Base (M€)`, `Revenue (M€)`, `Max foregone rev (M€)`, `Gain 25% mobil. (M€)`),
    decimals = 1
  ) %>%
  fmt_number(
    columns = c(`N articles`, `Base/article (€)`, `Struct. vacant dwell.`),
    decimals = 0
  ) %>%
  fmt_number(
    columns = c(`Rate commune (%)`, `Rate métropole (%)`, `Total rate (%)`, `Struct. vacancy rate`),
    decimals = 2
  ) %>%
  cols_label(
    Year                    = "Year",
    `Base (M€)`             = "Tax base (M€)",
    `Rate commune (%)`      = "Commune rate (%)",
    `Rate métropole (%)`    = "Metro rate (%)",
    `Total rate (%)`        = "Total rate (%)",
    `Revenue (M€)`          = "Revenue (M€)",
    `N articles`            = "Tax units",
    `Base/article (€)`      = "Base per unit (€)",
    `Struct. vacant dwell.` = "Structural vacant dwellings",
    `Struct. vacancy rate`  = "Structural vacancy rate (%)",
    `Max foregone rev (M€)` = "Foregone revenue (M€)",
    `Gain 25% mobil. (M€)`  = "Potential gain (M€)"
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#DBEAFE"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(rows = Year == max(Year))
  ) %>%
  tab_source_note(
    source_note = "Source: LOVAC open data, CEREMA / DGFiP"
  )
t
```
## Export Table
```{r}
gtsave(t, here("output", "LOVAC — Fiscal summary.png"))
```

# Integrated Analysis
## Revenue foregone vs actual
```{r}
g1 <- rei_lovac %>%
  select(annee, montant_reel, recettes_potentielles) %>%
  pivot_longer(-annee, names_to = "series", values_to = "value") %>%
  mutate(series = recode(series,
    montant_reel          = "Actual TFPB revenue",
    recettes_potentielles = "Max foregone (100% structural vacancies)"
  )) %>%
  ggplot(aes(x = annee, y = value / 1e6,
             color = series, linetype = series, group = series)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_color_manual(values = c("#1565C0", "#E74C3C")) +
  scale_x_continuous(breaks = 2020:2024) +
  labs(
    title    = "Actual TFPB revenue vs. maximum foregone revenue",
    subtitle = "Foregone = all structural vacancies (2+ yr) at current rate & base/article",
    x = NULL, y = "€ millions", color = NULL, linetype = NULL,
    caption = "Source: REI + LOVAC, DGFiP / CEREMA"
  )
g1
```

## Mobilization scenarios (stacked bar)
```{R}
g2 <- rei_lovac %>%
  select(annee, gain_10pct, gain_25pct, gain_50pct) %>%
  pivot_longer(-annee, names_to = "scenario", values_to = "gain") %>%
  mutate(scenario = recode(scenario,
    gain_10pct = "10% mobilized",
    gain_25pct = "25% mobilized",
    gain_50pct = "50% mobilized"
  )) %>%
  ggplot(aes(x = factor(annee), y = gain / 1e6, fill = scenario)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("#90CAF9", "#42A5F5", "#1565C0")) +
  labs(
    title    = "Potential TFPB gain from vacancy mobilization",
    subtitle = "By mobilization scenario (% of structural vacancies re-entering base)",
    x = NULL, y = "Additional revenue (M€)", fill = NULL,
    caption = "Source: REI + LOVAC, DGFiP / CEREMA"
  )
g2
```

## Rate hike vs vacancy mobilization (strategy comparison)
```{R}
# How many percentage points of rate increase would be needed to match
# each mobilization scenario?
g3 <- rei_lovac %>%
  mutate(
    # Equivalent rate hike (pp) to match each mobilization gain
    eq_rate_10 = gain_10pct / (base_nette / 100),
    eq_rate_25 = gain_25pct / (base_nette / 100),
    eq_rate_50 = gain_50pct / (base_nette / 100)
  ) %>%
  select(annee, eq_rate_10, eq_rate_25, eq_rate_50) %>%
  pivot_longer(-annee, names_to = "scenario", values_to = "equiv_rate_pp") %>%
  mutate(scenario = recode(scenario,
    eq_rate_10 = "10% vacancy mobil.",
    eq_rate_25 = "25% vacancy mobil.",
    eq_rate_50 = "50% vacancy mobil."
  )) %>%
  ggplot(aes(x = annee, y = equiv_rate_pp,
             color = scenario, group = scenario)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("#90CAF9", "#42A5F5", "#1565C0")) +
  scale_x_continuous(breaks = 2020:2024) +
  labs(
    title    = "Rate hike equivalent of each vacancy mobilization scenario",
    subtitle = "How many additional pp of TFPB rate would yield the same revenue?",
    x = NULL, y = "Equivalent rate increase (pp)", color = NULL,
    caption = "Source: REI + LOVAC"
  )
g3
```

## Structural vacancy rate vs base growth (dual axis)
```{R}
# Shows the tension: if vacant dwellings re-enter, base grows — and vice versa
g4_data <- rei_lovac %>%
  select(annee, tx_lv_sup2, croissance_base) %>%
  drop_na()
 
g4 <- ggplot(g4_data, aes(x = annee)) +
  geom_col(aes(y = tx_lv_sup2 * 100), fill = "#E74C3C", alpha = 0.4, width = 0.5) +
  geom_line(aes(y = croissance_base), color = "#1565C0", linewidth = 1.2) +
  geom_point(aes(y = croissance_base), color = "#1565C0", size = 3) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40") +
  scale_x_continuous(breaks = 2020:2024) +
  labs(
    title    = "Structural vacancy rate (bars) vs. tax base growth (line)",
    subtitle = "Blue line: annual base growth %; Red bars: structural vacancy rate %",
    x = NULL, y = "% ",
    caption = "Source: REI + LOVAC"
  )
g4
```

## Three-strategy revenue comparison (final decision chart)
```{R}
# For the most recent year: compare the three strategies side by side
latest <- rei_lovac %>% filter(annee == max(annee))
 
strategies <- tibble(
  strategy = c(
    "Do nothing",
    "Rate +1pp",
    "Rate +2pp",
    "25% vacancy\nmobilization",
    "50% vacancy\nmobilization",
    "Mixed: +1pp rate\n+ 25% mobil."
  ),
  gain_M = c(
    0,
    latest$gain_1pp_rate / 1e6,
    2 * latest$gain_1pp_rate / 1e6,
    latest$gain_25pct / 1e6,
    latest$gain_50pct / 1e6,
    (latest$gain_1pp_rate + latest$gain_25pct) / 1e6
  ),
  type = c("Baseline", "Rate", "Rate",
           "Base/vacancy", "Base/vacancy", "Mixed")
)
 
g5 <- strategies %>%
  mutate(strategy = fct_reorder(strategy, gain_M)) %>%
  ggplot(aes(x = strategy, y = gain_M, fill = type)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0("+", round(gain_M, 1), " M€")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Baseline"     = "grey80",
    "Rate"         = "#E74C3C",
    "Base/vacancy" = "#3498DB",
    "Mixed"        = "#8E44AD"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  labs(
    title    = "Revenue gain by strategy — Marseille",
    subtitle = paste0("Simulated for year ", max(rei_lovac$annee),
                      " using actual REI base & LOVAC vacancy stock"),
    x = NULL, y = "Additional annual revenue vs. do-nothing (M€)", fill = "Strategy type",
    caption = "Source: REI + LOVAC | Rate strategies = municipal rate only"
  )
g5
```

## Panel creation
```{r}
(g1 + g2) / (g3 + g4) / g5 +
  plot_annotation(
    title   = "Marseille TFPB — REI × LOVAC integrated analysis (2020–2024)",
    caption = "Source: REI (DGFiP) + LOVAC (CEREMA / DGFiP)"
  )
ggsave(here("output", "rei_lovac_integrated.png"),
       width = 14, height = 16, dpi = 150)
```